Smooth and Persistent Forecasts of German GDP:\\ Balancing Accuracy and Stability\thanks{We would like Barbara Rossi and the participants at the Vienna Forecasting Workshop 2025 for their comments and suggestions.}

### 2025-08-11 Archive IWH Simon paper/10.08.2025/paper.tex
up

\title{Smooth and Persistent Forecasts of German GDP\\ Balancing Accuracy and Stability\thanks{We would like Barbara Rossi and the participants at the Vienna Forecasting Workshop 2025 for their comments and suggestions.}}
\author{Katja Heinisch\thanks{Halle Institute for Economic Research (IWH).} \hspace{0.2cm}
Simon van Norden\thanks{HEC Montréal and CIREQ.} \hspace{0.2cm}
Marc Wildi\thanks{Zurich University of Applied Sciences (ZHAW).}}
\date{July 2025 \\
\vspace{2cm} \textit{Work in progress, not for circulation}}

@@

\begin{abstract} %max.
This paper explores the trade-off between informational efficiency and forecast smoothness in economic forecasting. Forecasts that minimize mean squared forecast errors often exhibit excessive volatility, limiting their practical applicability. We propose the Multivariate Smooth Sign Accuracy (M-SSA) framework as a solution, extracting smoothed components from leading indicators. This enhances the signal-to-noise ratio and controls forecast volatility and timing. Applied to German GDP, our method yields smoothed forecasts that improve the forecasting accuracy of traditional models, particularly over medium-term horizons. While smoother forecasts tend to lag slightly, this drawback can be mitigated by adjusting the forecast horizon. These findings highlight the practicality of the M-SSA framework for both forecasters and policymakers, especially in settings where updating forecasts or policy adjustments are costly.

@@

\bigskip \vspace{10pt}
\textbf{Keywords:} Forecast Smoothing, Smooth Sign Accuracy 
\newline
\textbf{JEL Classification:} C53, E37, E66
%C53 – Forecasting and Prediction Methods; Simulation Methods
%E37 – Forecasting and Simulation: Models and Applications
%E66 – General Outlook and Conditions: Forecasts and Simulations
\thispagestyle{empty}

@@

Economic forecasters face a dilemma
: while informational efficiency requires that forecasts incorporate all available data, doing so often results in forecasts that are excessively volatile.
 %Volatile forecasts may be a sign that they efficiently incorporate all available information. 
 In addition, forecasts may reflect excessive ``noise'' due to overfitting. Further, forecast users (including public and private decision makers) may face constraints in their speed of adjustment to changing forecasts. For example, fixed annual budgets may constrain spending, or transaction costs may limit changes in investments. Others may attempt to smooth their reactions for reasons of signalling. Important examples include dividend smoothing by corporate managers and interest rate smoothing by central banks.

@@

This study extends this framework by replacing raw leading indicators with smoothed and filtered components, thereby improving the signal-to-noise ratio in the predictor variables. We apply this framework to well-established German economic indicators to show its ability to improve forecasts of GDP \citep{lehmann2021predicting,Heinisch2018bottom}  and GDP trends using a variety of sample periods and performance metrics.

@@

We proceed as follows. Section \eqref{sec:data} describes our data series, the sample period used, and documents their dynamic behaviour. Section \eqref{sec:direct_forecast} provides benchmark forecasts using simple direct projections of GDP based on selected indicators. We demonstrate how in-sample forecast performance may improve when we ``de-noise'' the indicators prior to projection with simple smoothing filters. Section \eqref{sec:mSSA} defines the M-SSA method and shows how its forecast performance compares to that of direct projections as well as how it varies with forecast horizon. 
Section \eqref{sec:mSSA_component} considers the ability of our indicators to forecast German GDP growth out-of-sample using the M-SSA framework, and Section \eqref{sec:revisions} investigates the temporal stability of the estimated forecasting equations. 
Section \eqref{sec:conclusions} summarizes our results and concludes.

\section{Data and Dependence}\label{sec:data}
\subsection{Data and Publication Lags}
The analysis employs a set of well-established leading indicators to forecast quarterly German GDP \citep{Drechsel2012financial,Heinisch2018bottom}, including industrial production (IP), the ifo Business Climate Index (ifo\_c), the Economic Sentiment Indicator (ESI), and the term spread between 10-year bond yields and the 3-month EURIBOR rate (spre\_10y\_3m) covering the period 1995 to 2024.  
We take into account the publication lag of the indicators and the GDP series (Table \eqref{tab:data_stucture}).\footnote{Data revisions are not considered, given that previous studies \citep{Heinisch2019} have shown that their overall effect on forecast performance is minor.} 
While GDP is available with a one-quarter delay, and IP is released with a two-month lag, business surveys and financial data are published without any lag. %All series are consequently re-aligned to correspond with the information available to forecasters, with GDP and IP variables shifted forward according to their respective lags. Table \eqref{tab:data_stucture_aligned} displays the dataset as of the end of 2024, with GDP and IP variables temporally aligned at the sample endpoint. This configuration reflects the matrix of explanatory variables accessible in Q4-2024 for the purposes of nowcasting or forecasting GDP. For illustrative purposes, an additional target column is included on the right side of the table, representing the dependent variable for nowcasting applications. In the case of an $h$-step ahead forecast, this column is shifted upwards by $h$ quarters.

\begin{table}[ht]
\centering
\begin{tabular}{rrrrrr}
  \hline
 & GDP & IP & ifo\_c & ESI & spr\_10y\_3m \\ 
  \hline
2024-07-01 &  & 90.700 & 85.800 & 92.500 & -1.300 \\ 
  2024-08-01 &  & 93.200 & 84.900 & 90.900 & -1.400 \\ 
  2024-09-01 &902.571  & 91.200 & 83.700 & 89.800 & -1.400 \\ 
  2024-10-01 &  & 90.800 & 84.100 & 90.700 & -1.200 \\ 
  2024-11-01 &  & 92.200 & 83.800 & 89.300 & -0.900 \\ 
  2024-12-01 &  &  & 82.800 & 86.900 & -0.800 \\ 
  2025-01-01 &  &  & 82.500 & 88.100 & -0.200 \\ 
   \hline
\end{tabular}
\caption{Ragged edge structure. As of January 2025, before publication of the GDP flash estimate.} 
\label{tab:data_stucture}
\end{table}

To allow for missing observations due to publication lags, we re-align series to reflect the information available to forecasters. Table \eqref{tab:data_stucture_aligned} displays the result as of the end of 2024, with GDP and IP columns shifted to reflect their release dates rather than the periods that they measure. This will be the data configuration used for nowcasting or forecasting GDP. For illustrative purposes, an additional target column is included on the right side of the table, representing the dependent variable for nowcasting applications.\footnote{In the case of an $h$-step ahead forecast, this column is shifted upwards by $h$ quarters.}

\begin{table}[ht]
\centering
\begin{tabular}{rrrrrrrc}
  \hline
 & GDP & IP & ifo\_c & ESI & spr\_10y\_3m &&Target (nowcast)\\ 
  \hline
  2024-07-01 & & 93.400 & 85.800 & 92.500 & -1.300&|& \\ 
  2024-08-01 & & 90.700&  84.900 & 90.900 & -1.400&| &\\ 
  2024-09-01 & 901.623 &93.200 & 83.700 & 89.800 & -1.400&| &902.571\\ 
  2024-10-01 &  &91.200 & 84.100 & 90.700 & -1.200&| &\\ 
  2024-11-01 &  & 90.800& 83.800 & 89.300 & -0.900&| &\\ 
  2024-12-01 &902.571   & 92.200  & 82.800 & 86.900 & -0.800&| &?\\ 
    \hline
\end{tabular}
\caption{Aligned dataset. At each date, columns 1 through 5 show the most recent values of each predictor available to forecasters. Column 6 represents the target variable for a GDP nowcast (i.e. the values of column 1 shifted upwards by the publication lag of GDP.} 
\label{tab:data_stucture_aligned}
\end{table}

To ensure comparability and mitigate scale effects, all indicators (except the term spread, which is differenced only) are log-differenced and standardized. Additionally, extreme outliers associated with the COVID-19 pandemic are trimmed to prevent distortions in subsequent figures. 
 Monthly series are shifted (Table \eqref{tab:data_stucture_aligned}) and averaged to quarterly series (Figure \eqref{fig:data}), while Figure \eqref{fig:data_lags} offers a close-up view of the financial crisis and the pandemic. The latter figure shows that the real-time GDP and industrial production series are synchronized at turning points, whereas the other indicators tend to lead somewhat. The challenge for forecasters is to use the latter in a multivariate approach to improve forecasting performance over that of univariate benchmarks.

@@

\subsection{Dependence}
To examine the joint dynamics of our variables, we estimate their cross-correlations as well as their vector moving-average (VMA) representations. The latter is a key input to the multivariate filters we develop in Sections \eqref{sec:mSSA} and \eqref{sec:mSSA_component}.

Figures \eqref{CCF} (full sample) and \eqref{CCF_wc} (omitting the Pandemic period) show the sample cross-correlation functions (CCF's) between current and lagged GDP and the two survey indicators. Both figures show strong correlations that peak when GDP is slightly lagged. Results for IP showed no such lag, while correlations for the term spread were weaker. The figures also show that the precise strengths of the correlations are influenced by the outliers associated with the COVID-19 pandemic. In the interest of model stability, we exclude data from Q4-2019 to Q4-2020 (pandemic) when modelling the dynamics for the multivariate filters developed below.

@@

We then estimated a variety of parsimonious VARMA models to capture these dynamics, selecting parsimonious VAR(1) and VAR(3) specifications.\footnote{We used the \texttt{refVARMA()} procedure in the \cite{MTSpackage} \texttt{MTS} package, which sets VARMA coefficients with t-statistics below a critical threshold to zero \citep[see][]{tsay2013multivariate}.  This led to a VAR(1) specification with a threshold of 1.5. The VAR(1) left some residual evidence of higher-order dynamics, but capturing these in our limited sample size was difficult without overfitting. We therefore explored VAR($p$) models with coefficients estimated by Elastic Net using the \cite{ElasticNet} \texttt{elasticnet} package.  This led to a VAR(3) specification with 7 steps. For additional details, see \cite{ElasticNet} and the references therein.} The figures \eqref{ma_inv_BIP} and \eqref{ma_inv_BIP_var3_en} show the implied responses of GDP to shocks to each indicator variable.\footnote{Shown are the MA coefficients for GDP from the vector moving-average (VMA) models obtained by inversion of the VAR(1) and VAR(3) representations respectively (i.e. the impulse-response functions (IRFs) of GDP to the correlated reduced-form shocks.)}

@@

Figures \eqref{direct_wc} and \eqref{direct_wc_financial_crisis} compare actual and fitted values for GDP at forecast horizons $h = 0, \ldots, 3$, using the four indicators and the target, i.e., GDP shifted $h$ quarters ahead. Figure \eqref{direct_wc} shows results over the sample without the pandemic, while Figure \eqref{direct_wc_financial_crisis} provides a closer look at the fit during the financial crisis. Table \eqref{tab:pvaluedhp} shows p-values for tests of the joint null hypothesis that all coefficients on the indicators are zero (HAC-adjusted Wald test).

@@

The results indicate that as the forecast horizon $h$ increases, forecast become less volatile and their turning points (i.e., peaks and troughs) become increasingly right-shifted (lagged) relative to the target GDP. Additionally, the p-values of the HAC-adjusted Wald test precipitously increase from $h=2$ to $h=3$ (see Table \eqref{tab:pvaluedhp}), implying that the indicators become statistically insignificant for forecast horizons of more than two quarters.\footnote{ 
These findings are confirmed by the relative root mean-square errors comparing direct forecasts against the mean benchmark, see Table \eqref{rRMSE_mSSA_direct_mean_without_covid8} in the Appendix.}

@@

%\begin{table}[ht]
%\centering
%\begin{tabular}{rrrrr}
%  \hline
% & h=3 & h=4 & h=5 & h=6 \\ 
%  \hline
%rRMSE & 0.962 & 0.967 & 0.990 & 0.967 \\ 
%   \hline
%\end{tabular}
%\caption{Relative root mean-square error of HP-C against classic direct forecast for $h\geq %3$.\\Values $<1$ imply  HP-C has smaller RMSE.  } 
%\label{tab:rRMSE}
%\end{table}
\bigskip
\begin{table}[h]
\centering
\begin{tabular}{lrrrrrr}
  \hline
  & h= 1 & h= 2 & h= 3 & h= 4 & h= 5 & h= 6 \\ 
  \hline
  Unfiltered & 0.000 & 0.001 & 0.910 & 0.509 & 0.091 & 0.178 \\ 
HP-C filtered & 0.001 & 0.036 & 0.046 & 0.049 & 0.012 & 0.022 \\ 
   \hline
\end{tabular}
\caption{p-values for $H_0: \boldsymbol{\beta} = 0$ in \eqref{eq:direct_forecast} based on HAC-adjusted Wald test.\\Estimation over full sample (without pandemic.)} 
\label{tab:pvaluedhp}
\end{table}

@@

Figure \eqref{hp_160} presents both classic two-sided and one-sided HP(160) (denoted HP-C for \textit{concurrent}) filters (top panels) along with their associated amplitude functions in the frequency domain (bottom panels). It confirms that the HP-C filter assigns greater weight to yearly components—relevant in a medium-term forecast context— than the classic quarterly HP filter, while also reducing undesirable high-frequency noise. Therefore, while the choice of the HP(160) filter in this example
is somewhat arbitrary, it should serve the purpose of reducing high-frequency noise while preserving longer-duration movements.\footnote{A comprehensive technical analysis of the effect of $\lambda$ on the resulting GDP predictor can be conducted using the M-SSA package (\url{https://github.com/wiaidp/R-package-SSA-Predictor}).}

@@

Figures \eqref{direct_hp_forecasts} and \eqref{direct_hp_forecasts_financial_crisis} provide results comparable to those of Figures \eqref{direct_wc} and \eqref{direct_wc_financial_crisis} but now using indicators smoothed with the one-sided HP(160), (denoted as the HP-C, for \textit{concurrent}) filter.  The fitted values lag GDP slightly less and track turning points somewhat more quickly. Table \eqref{tab:pvaluedhp} shows p-values for tests of the joint null hypothesis that all coefficients ${\boldsymbol{\beta}} = 0$. The use of filtered indicators improves their statistical significance at forecast horizons $h > 2$ quarters. 
%Figures \eqref{direct_hp_forecasts} (all observations with exclusion of the pandemic) and \eqref{direct_hp_forecasts_financial_crisis} (financial crisis). A comparison with the classic (unfiltered) direct forecasts in the latter figure %in Fig.\eqref{direct_wc_financial_crisis} 
% indicates that the new predictors are slightly less retarded and track the turning points more promptly (`faster'). 
% The filtered approach also exhibits superior statistical properties. As shown in Table \eqref{tab:f_stats}, 
%The p-values of the F-statistics for the HP-based forecasts are smaller than for the direct forecasts specifically at $h=3$, but the predictor remains statistically insignificant at forecast horizons exceeding two quarters (even though the evaluation is conducted in-sample and based on all available indicators).

@@

Let $\mathbf{x}_t$ be a time series representing the data, with observations $(x_{t},...,x_{1})'$. Let $\mathbf{z}_t=(z_{t},...,z_1)'$  denote a target series, which depends on $\mathbf{x}_t$ as well as on unknown future $x_{t+k}$, $k>0$ (and possibly unknown past $x_k$, $k\leq 0$). For simplicity, we assume stationarity of all time series involved. In our current framwork, $z_t$ corresponds to the output of a two-sided Hodrick–Prescott filter with smoothing parameter $\lambda = 160$ applied to the doubly infinite sequence $x_k$, $k\in\mathbb{Z}$. The finite sample $\mathbf{x}_t$ is the portion effectively observed and represents the log-differenced GDP (the dependence on $x_0, x_{-1},...$ is neglected because the HP filter weights decay rapidly to zero; see Fig. \eqref{hp_160}). The prediction task comprises tracking $z_{t}$  using the (causal) predictor $y_{t}=\sum_{k=0}^{L-1} b_{k}x_{t-k}=\mathbf{b}'\mathbf{x}_t$, where the filter $\mathbf{b}$ with weights $b_0,...,b_{L-1}'$ (actually of length $L=t$) is available for optimization purposes.

@@

Traditional predictors often yield an excessive number of false sign changes due to leakage of high-frequency components (high-frequency leakage) in the predicted series, as discussed above (see Figure \eqref{hp_160}). This noisiness is particularly problematic near the onset or termination of economic crises (recessions), where the target fluctuates around zero; reliable tracking of sign changes would facilitate real-time assessment of the occurrence and timing of such events. 
%as illustrated by the amplitude function in Figure \eqref{hp_160} (bottom right panel) and further elaborated upon by \cite{Wildi2025}. 
 \cite{Wildi2025} shows that the SSA criterion helps control this phenomenon (see Fig.\eqref{mssa_msse_zc} for illustration). Using the dual formulation of the criterion \eqref{critssaht}, the author demonstrates that the SSA solution yields the \emph{lowest} zero-crossing rate among all (linear) predictors with equivalent target correlation.
The concept can be extended to a multivariate framework, denoted M-SSA, by generalizing the correlation functions as detailed in the next section.\\

>>>>

\title{Smooth and Persistent Forecasts of German GDP:\\ Balancing Accuracy and Stability\thanks{We would like Barbara Rossi and the participants at the Vienna Forecasting Workshop 2025 for their comments and suggestions.}}
\author{Katja Heinisch\thanks{Halle Institute for Economic Research (IWH).} \hspace{0.2cm}
Simon van Norden\thanks{HEC Montréal and CIREQ.} \hspace{0.2cm}
Marc Wildi\thanks{Zurich University of Applied Sciences (ZHAW).}}
\date{August 2025 \\
\vspace{2cm} \textit{Work in progress, not for circulation}}

@@

\begin{abstract} %max.
This paper explores the trade-off between informational efficiency and forecast smoothness in economic forecasting. Forecasts that minimize mean squared forecast errors often exhibit excessive volatility, limiting their practical applicability. We propose the Multivariate Smooth Sign Accuracy (M-SSA) framework as a solution, extracting smoothed components from leading indicators. This enhances the signal-to-noise ratio and controls forecast volatility and timing. Applied to German GDP growth, our method yields smoothed forecasts that improve the forecasting accuracy of traditional models, particularly over medium-term horizons. While smoother forecasts tend to lag slightly, this drawback can be mitigated by adjusting the forecast horizon. These findings highlight the practicality of the M-SSA framework for both forecasters and policymakers, especially in settings where updating forecasts or policy adjustments are costly.

@@

\bigskip \vspace{10pt}
\textbf{Keywords:} Forecast Smoothing, Smooth Sign Accuracy, Time Series Filtering 
\newline
\textbf{JEL Classification:} C53, E37, E66
%C53 – Forecasting and Prediction Methods; Simulation Methods
%E37 – Forecasting and Simulation: Models and Applications
%E66 – General Outlook and Conditions: Forecasts and Simulations
\thispagestyle{empty}

@@

Economic forecasters face a dilemma: while informational efficiency requires that forecasts incorporate all available data, doing so often results in forecasts that are excessively volatile.
 %Volatile forecasts may be a sign that they efficiently incorporate all available information. 
 In addition, forecasts may reflect excessive ``noise'' due to overfitting. Further, forecast users (including public and private decision makers) may face constraints in their speed of adjustment to changing forecasts. For example, fixed annual budgets may constrain spending, or transaction costs may limit changes in investments. Others may attempt to smooth their reactions for reasons of signalling. Important examples include dividend smoothing by corporate managers and interest rate smoothing by central banks.

@@

This study extends this framework by replacing raw leading indicators with smoothed and filtered components, thereby improving the signal-to-noise ratio in the predictor variables. We apply this framework to well-established German economic indicators to show its ability to improve forecasts of GDP growth \citep{lehmann2021predicting,Heinisch2018bottom}  and GDP trends using a variety of sample periods and performance metrics.

@@

We proceed as follows. Section \ref{sec:data} describes our data series, the sample period used, and documents their dynamic behaviour. Section \ref{sec:direct_forecast} provides benchmark forecasts using simple direct projections of GDP based on selected indicators. We demonstrate how in-sample forecast performance may improve when we ``de-noise'' the indicators prior to projection with simple smoothing filters. Section \ref{sec:mSSA} defines the M-SSA method and shows how its forecast performance compares to that of direct projections as well as how it varies with forecast horizon. 
Section \ref{sec:mSSA_component} considers the ability of our indicators to forecast German GDP growth out-of-sample using the M-SSA framework, and Section \ref{sec:revisions} investigates the temporal stability of the estimated forecasting equations. Section \ref{sec:conclusions} summarizes our results and concludes.

\section{Data and Dependence}\label{sec:data}
\subsection{Data and Publication Lags}
The analysis employs a set of well-established leading indicators to forecast quarterly German GDP \citep{Drechsel2012financial,Heinisch2018bottom}, including industrial production (IP), the ifo Business Climate Index (ifo\_c), the Economic Sentiment Indicator (ESI), and the term spread between 10-year bond yields and the 3-month EURIBOR rate (spre\_10y\_3m) covering the period 1995 to 2024.  
We take into account the publication lag of the indicators and the GDP series (Table \ref{tab:data_stucture}).\footnote{Data revisions are not considered, given that previous studies \citep{Heinisch2019} have shown that their overall effect on forecast performance is minor.} 
While GDP is available with a one-quarter delay, and IP is released with a two-month lag, business surveys and financial data are published without any lag. %All series are consequently re-aligned to correspond with the information available to forecasters, with GDP and IP variables shifted forward according to their respective lags. Table \eqref{tab:data_stucture_aligned} displays the dataset as of the end of 2024, with GDP and IP variables temporally aligned at the sample endpoint. This configuration reflects the matrix of explanatory variables accessible in Q4-2024 for the purposes of nowcasting or forecasting GDP. For illustrative purposes, an additional target column is included on the right side of the table, representing the dependent variable for nowcasting applications. In the case of an $h$-step ahead forecast, this column is shifted upwards by $h$ quarters.

\begin{table}[ht]
\caption{Ragged edge structure. As of January 2025, before publication of the GDP flash estimate.} 
\label{tab:data_stucture}
\centering
\begin{tabular}{rrrrrr}
  \hline
 & GDP & IP & ifo\_c & ESI & spr\_10y\_3m \\ 
  \hline
2024-07-01 &  & 90.700 & 85.800 & 92.500 & -1.300 \\ 
  2024-08-01 &  & 93.200 & 84.900 & 90.900 & -1.400 \\ 
  2024-09-01 &902.571  & 91.200 & 83.700 & 89.800 & -1.400 \\ 
  2024-10-01 &  & 90.800 & 84.100 & 90.700 & -1.200 \\ 
  2024-11-01 &  & 92.200 & 83.800 & 89.300 & -0.900 \\ 
  2024-12-01 &  &  & 82.800 & 86.900 & -0.800 \\ 
  2025-01-01 &  &  & 82.500 & 88.100 & -0.200 \\ 
   \hline
\end{tabular}

\end{table}

To allow for missing observations due to publication lags, we re-align series to reflect the information available to forecasters. Table \ref{tab:data_stucture_aligned} displays the artificial result as of the end of 2024, with GDP and IP columns shifted to reflect their release dates rather than the periods that they measure. This data configuration will be the used for nowcasting or forecasting GDP. For illustrative purposes, an additional target column is included on the right side of the table, representing the dependent variable for nowcasting applications.\footnote{In the case of an $h$-step ahead forecast, this column is shifted upwards by $h$ quarters.}

\begin{table}[ht]
\caption{Aligned dataset. At each date, various columns show the most recent values of each predictor available to forecasters. Column 6 represents the target variable for a GDP nowcast (i.e. the values of column 1 shifted upwards by the publication lag of GDP.} 
\label{tab:data_stucture_aligned}
\centering
\begin{tabular}{rrrrrrrc}
  \hline
 & GDP & IP & ifo\_c & ESI & spr\_10y\_3m &&Target (nowcast)\\ 
  \hline
  2024-07-01 & & 93.400 & 85.800 & 92.500 & -1.300&|& \\ 
  2024-08-01 & & 90.700&  84.900 & 90.900 & -1.400&| &\\ 
  2024-09-01 & 901.623 &93.200 & 83.700 & 89.800 & -1.400&| &902.571\\ 
  2024-10-01 &  &91.200 & 84.100 & 90.700 & -1.200&| &\\ 
  2024-11-01 &  & 90.800& 83.800 & 89.300 & -0.900&| &\\ 
  2024-12-01 &902.571   & 92.200  & 82.800 & 86.900 & -0.800&| &?\\ 
    \hline
\end{tabular}

\end{table}

To ensure comparability and mitigate scale effects, all indicators (except the term spread, which is differenced only) are log-differenced and standardized. Additionally, extreme outliers associated with the COVID-19 pandemic are trimmed to prevent distortions in subsequent figures. 
 Monthly series are shifted (Table \ref{tab:data_stucture_aligned}) and averaged to quarterly series (Figure \ref{fig:data}), while Figure \ref{fig:data_lags} offers a close-up view of the financial crisis and the pandemic. The latter figure shows that the real-time GDP and industrial production series are synchronized at turning points, whereas the other indicators tend to lead somewhat. The challenge for forecasters is to use the latter in a multivariate approach to improve forecasting performance over that of univariate benchmarks.

@@

\subsection{Dependence}
To examine the joint dynamics of our variables, we estimate their cross-correlations as well as their vector moving-average (VMA) representations. The latter is a key input to the multivariate filters we develop in Sections \ref{sec:mSSA} and \ref{sec:mSSA_component}.

Figures \ref{CCF} (full sample) and \ref{CCF_wc} (omitting the Pandemic period) show the sample cross-correlation functions (CCF's) between current and lagged GDP and the two survey indicators. Both figures show strong correlations that peak when GDP is slightly lagged. Results for IP showed no such lag, while correlations for the term spread were weaker. The figures also show that the precise strengths of the correlations are influenced by the outliers associated with the COVID-19 pandemic. In the interest of model stability, we exclude data from Q4-2019 to Q4-2020 (pandemic) when modelling the dynamics for the multivariate filters developed below.

@@

We then estimated a variety of parsimonious VARMA models to capture these dynamics, selecting parsimonious VAR(1) and VAR(3) specifications.\footnote{We used the \texttt{refVARMA()} procedure in the \cite{MTSpackage} \texttt{MTS} package, which sets VARMA coefficients with t-statistics below a critical threshold to zero \citep[see][]{tsay2013multivariate}.  This led to a VAR(1) specification with a threshold of 1.5. The VAR(1) left some residual evidence of higher-order dynamics, but capturing these in our limited sample size was difficult without overfitting. We therefore explored VAR($p$) models with coefficients estimated by Elastic Net using the \cite{ElasticNet} \texttt{elasticnet} package.  This led to a VAR(3) specification with 7 steps. For additional details, see \cite{ElasticNet} and the references therein.} The figures \ref{ma_inv_BIP} and \ref{ma_inv_BIP_var3_en} show the implied responses of GDP to shocks to each indicator variable.\footnote{Shown are the MA coefficients for GDP from the vector moving-average (VMA) models obtained by inversion of the VAR(1) and VAR(3) representations respectively (i.e. the impulse-response functions (IRFs) of GDP to the correlated reduced-form shocks.)}

@@

Figures \ref{direct_wc} and \ref{direct_wc_financial_crisis} compare actual and fitted values for GDP at forecast horizons $h = 0, \ldots, 3$, using the four indicators and the target, i.e., GDP shifted $h$ quarters ahead. Figure \ref{direct_wc} shows results over the sample without the pandemic, while Figure \ref{direct_wc_financial_crisis} provides a closer look at the fit during the financial crisis. Table \ref{tab:pvaluedhp} shows p-values for tests of the joint null hypothesis that all coefficients on the indicators are zero (HAC-adjusted Wald test).

@@

The results indicate that as the forecast horizon $h$ increases, forecast become less volatile and their turning points (i.e., peaks and troughs) become increasingly right-shifted (lagged) relative to the target GDP. Additionally, the p-values of the HAC-adjusted Wald test precipitously increase from $h=2$ to $h=3$ (see Table \eqref{tab:pvaluedhp}), implying that the indicators become statistically insignificant for forecast horizons of more than two quarters.\footnote{ 
These findings are confirmed by the relative root mean-square errors comparing direct forecasts against the mean benchmark, see Table \ref{rRMSE_mSSA_direct_mean_without_covid8} in the Appendix.}

@@

%\begin{table}[ht]
%\centering
%\begin{tabular}{rrrrr}
%  \hline
% & h=3 & h=4 & h=5 & h=6 \\ 
%  \hline
%rRMSE & 0.962 & 0.967 & 0.990 & 0.967 \\ 
%   \hline
%\end{tabular}
%\caption{Relative root mean-square error of HP-C against classic direct forecast for $h\geq %3$.\\Values $<1$ imply  HP-C has smaller RMSE.  } 
%\label{tab:rRMSE}
%\end{table}
\bigskip
\begin{table}[h]
\caption{p-values for $H_0: \boldsymbol{\beta} = 0$ in \eqref{eq:direct_forecast} based on HAC-adjusted Wald test.\\Estimation over full sample (without pandemic.)} 
\label{tab:pvaluedhp}
\centering
\begin{tabular}{lrrrrrr}
  \hline
  & h= 1 & h= 2 & h= 3 & h= 4 & h= 5 & h= 6 \\ 
  \hline
  Unfiltered & 0.000 & 0.001 & 0.910 & 0.509 & 0.091 & 0.178 \\ 
HP-C filtered & 0.001 & 0.036 & 0.046 & 0.049 & 0.012 & 0.022 \\ 
   \hline
\end{tabular}

\end{table}

@@

Figure \ref{hp_160} presents both classic two-sided and one-sided HP(160) (denoted HP-C for \textit{concurrent}) filters (top panels) along with their associated amplitude functions in the frequency domain (bottom panels). It confirms that the HP-C filter assigns greater weight to yearly components—relevant in a medium-term forecast context— than the classic quarterly HP filter, while also reducing undesirable high-frequency noise. Therefore, while the choice of the HP(160) filter in this example
is somewhat arbitrary, it should serve the purpose of reducing high-frequency noise while preserving longer-duration movements.\footnote{A comprehensive technical analysis of the effect of $\lambda$ on the resulting GDP predictor can be conducted using the M-SSA package (\url{https://github.com/wiaidp/R-package-SSA-Predictor}).}

@@

Figures \ref{direct_hp_forecasts} and \ref{direct_hp_forecasts_financial_crisis} provide results comparable to those of Figures \ref{direct_wc} and \ref{direct_wc_financial_crisis} but now using indicators smoothed with the one-sided HP(160), (denoted as the HP-C, for \textit{concurrent}) filter.  The fitted values lag GDP slightly less and track turning points somewhat more quickly. Table \ref{tab:pvaluedhp} shows p-values for tests of the joint null hypothesis that all coefficients ${\boldsymbol{\beta}} = 0$. The use of filtered indicators improves their statistical significance at forecast horizons $h > 2$ quarters. 
%Figures \ref{direct_hp_forecasts} (all observations with exclusion of the pandemic) and \ref{direct_hp_forecasts_financial_crisis} (financial crisis). A comparison with the classic (unfiltered) direct forecasts in the latter figure %in Fig.\eqref{direct_wc_financial_crisis} 
% indicates that the new predictors are slightly less retarded and track the turning points more promptly (`faster'). 
% The filtered approach also exhibits superior statistical properties. As shown in Table \ref{tab:f_stats}, 
%The p-values of the F-statistics for the HP-based forecasts are smaller than for the direct forecasts specifically at $h=3$, but the predictor remains statistically insignificant at forecast horizons exceeding two quarters (even though the evaluation is conducted in-sample and based on all available indicators).

@@

Let $\mathbf{x}_t$ be a time series representing the data, with observations $(x_{t},...,x_{1})'$. Let $\mathbf{z}_t=(z_{t},...,z_1)'$  denote a target series, which depends on $\mathbf{x}_t$ as well as on unknown future $x_{t+k}$, $k>0$ (and possibly unknown past $x_k$, $k\leq 0$). For simplicity, we assume stationarity of all time series involved. In our current framwork, $z_t$ corresponds to the output of a two-sided Hodrick–Prescott filter with smoothing parameter $\lambda = 160$ applied to the doubly infinite sequence $x_k$, $k\in\mathbb{Z}$. The finite sample $\mathbf{x}_t$ is the portion effectively observed and represents the log-differenced GDP (the dependence on $x_0, x_{-1},...$ is neglected because the HP filter weights decay rapidly to zero; see Fig. \ref{hp_160}). The prediction task comprises tracking $z_{t}$  using the (causal) predictor $y_{t}=\sum_{k=0}^{L-1} b_{k}x_{t-k}=\mathbf{b}'\mathbf{x}_t$, where the filter $\mathbf{b}$ with weights $b_0,...,b_{L-1}'$ (actually of length $L=t$) is available for optimization purposes.

@@

Traditional predictors often yield an excessive number of false sign changes due to leakage of high-frequency components (high-frequency leakage) in the predicted series, as discussed above (see Figure \ref{hp_160}). This noisiness is particularly problematic near the onset or termination of economic crises (recessions), where the target fluctuates around zero; reliable tracking of sign changes would facilitate real-time assessment of the occurrence and timing of such events. 
%as illustrated by the amplitude function in Figure \ref{hp_160} (bottom right panel) and further elaborated upon by \cite{Wildi2025}. 
 \cite{Wildi2025} shows that the SSA criterion helps control this phenomenon (see Fig.\ref{mssa_msse_zc} for illustration).\\ Using the dual formulation of the criterion \eqref{critssaht}, the author demonstrates that the SSA solution yields the \emph{lowest} zero-crossing rate among all (linear) predictors with equivalent target correlation. 
The concept can be extended to a multivariate framework, denoted M-SSA, by generalizing the correlation functions as detailed in the next section. As we shall see, controlling the zero-crossing rate within the M-SSA framework will have a quantifiable effect on the hit rate and false positive rate, thereby explicitly addressing the detection of present (nowcasting) and future (forecasting) economic contractions and expansions.\\
